An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-05-02 cal_demo/ms.tex
describe F-engine and FX in cal paper

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a 
traditional FX or XF correlator, pair-wise visibilities are written to disk and used to calibrate 
offline before further processing such as imaging. However, a direct imaging correlator mixes 
the signals from all antennas before averaging and writing to disk, making calibration a 
requirement at the front end, before imaging and averaging. Previous solutions have involved 
applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or 
integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve 
for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of 
FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

Next we turn to the measurement performed by an antenna with far-field radiation pattern 
$\mathcal{W}_a(\hat{\mathbf{s}})$, where the subscript $a$ denotes a specific antenna. We 
adopt the common convention of normalizing the radiation pattern to have response of 1 at the 
center of the main lobe \citep[e.g.][]{nap99}. Consequently, the translation between the terminal 
voltages to electric field measurements has been absorbed in our definition of the radiation 
pattern. The ideal antenna electric field measurement can be expressed as 
\begin{equation}
E^I_a(f,t) = \int \mathcal{W}_a(\hat{\mathbf{s}})\mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega. 
\end{equation}
Equivalently we can express the effect of the antenna radiation pattern in the plane of the 
aperture.
\begin{equation}
E^I_a(f,t) = \int_{\mathrm{aperture}}W_a(\mathbf{r}-\mathbf{r}_a)E_a(\mathbf{r},t)\,d^2\mathbf{r}
\end{equation}
Here we used the Fourier relationship between the antenna's far-field radiation pattern and the 
aperture distribution, $W_a(\mathbf{r}-\mathbf{r}_a)=\int  \mathcal{W}_a(\hat{\mathbf{s}}) \, d
\Omega$. Although any antenna extends over a region of space, we include a nominal antenna 
position, $\mathbf{r}_a$, for multiple reasons. In the case where all antennas of an array are 
identical, a global aperture distribution can be adopted and the translation of the antennas 
distinguishes them. However, even if all antennas have unique aperture distributions, it can be 
convenient to retain the relative positions of antennas in the analysis. This allows for easy 
comparison with visibility-based analyses where the baseline vectors are used to describe 
sampling of the $uv$ plane. From a practical point of view, measurements of the antenna 
radiation pattern may not be performed in-situ, so it is convenient to describe the antenna 
aperture distribution relative to a point fixed within the aperture, along with a translational 
position.

@@

The first step of the MOFF algorithm is identical to the F-engine of a traditional FX correlator 
where the electric fields are Fourier transformed to frequency channels, $\Er{a}^T(f,t)$. Even 
after transforming along the time axis, we retain the time dependence of the signal to represent 
a stream of spectra.

>>>>

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a 
traditional cross correlator, pair-wise visibilities are written to disk and used to calibrate 
offline before further processing such as imaging. However, a direct imaging correlator mixes 
the signals from all antennas before averaging and writing to disk, making calibration a 
requirement at the front end, before imaging and averaging. Previous solutions have involved 
applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or 
integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve 
for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of 
FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

Next we turn to the measurement performed by an antenna with far-field radiation pattern 
$\mathcal{W}_a(\hat{\mathbf{s}})$, where the subscript $a$ denotes a specific antenna. We 
adopt the common convention of normalizing the radiation pattern to have response of 1 at the 
center of the main lobe \citep[e.g.][]{nap99}. Consequently, the translation between the terminal 
voltages to electric field measurements has been absorbed in our definition of the radiation 
pattern. The ideal antenna electric field measurement can be expressed as 
\begin{equation}
E^I_a(f,t) = \int \mathcal{W}_a(\hat{\mathbf{s}})\mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega. 
\end{equation}
Equivalently we can express the effect of the antenna radiation pattern in the plane of the 
aperture.
\begin{equation}
E^I_a(f,t) = \int_{\mathrm{aperture}}W_a(\mathbf{r}-\mathbf{r}_a)E_a(\mathbf{r},f,t)\,d^2\mathbf{r}
\end{equation}
Here we used the Fourier relationship between the antenna's far-field radiation pattern and the 
aperture distribution, $W_a(\mathbf{r}-\mathbf{r}_a)=\int  \mathcal{W}_a(\hat{\mathbf{s}}) \, d
\Omega$. Although any antenna extends over a region of space, we include a nominal antenna 
position, $\mathbf{r}_a$, for multiple reasons. In the case where all antennas of an array are 
identical, a global aperture distribution can be adopted and the translation of the antennas 
distinguishes them. However, even if all antennas have unique aperture distributions, it can be 
convenient to retain the relative positions of antennas in the analysis. This allows for easy 
comparison with visibility-based analyses where the baseline vectors are used to describe 
sampling of the $uv$ plane. From a practical point of view, measurements of the antenna 
radiation pattern may not be performed in-situ, so it is convenient to describe the antenna 
aperture distribution relative to a point fixed within the aperture, along with a translational 
position.

@@

The first step of the MOFF algorithm is to digitize the antenna signals and perform an FFT to 
channelize the time-ordered data. This is effectively estimating the quasi-monochromatic 
components of the electric field we already discussed. The sampling rate of the digitization 
determines the bandwidth of the observation, and the sampling period determines the minimum 
frequency channel resolution. The frequency Fourier transform operation is often referred to as 
the F-engine of a correlator. The MOFF shares the same F-engine design as an FX cross-
correlator which performs the frequency transform (F-engine) before cross multiplying (X-
engine).

%The first step of the MOFF algorithm is identical to the F-engine of a traditional FX correlator 
%where the electric fields are Fourier transformed to frequency channels, $\Er{a}^T(f,t)$. Even 
%after transforming along the time axis, we retain the time dependence of the signal to represent 
%a stream of spectra.
